---
title: PS270 Final Project
author: "Mallory Bottemiller"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

Does a lake having public boat access effect the likely hood of a zebra mussel presence, and if it does what type of landing influences more zebra mussel presence? My hypothesis is that lakes with public boat access causes zebra mussel presence. This hypothesis is based on the fact that people are more likely to deteriorate public goods like public lakes, applying this theory people would not care to check their equipment for invasive species before using it in the common good. My outcome variable in this case is the presence of zebra mussels, which can be measured by presence in a lake. My explanatory variable is if a lake has public boat access or not, and if it does what type of landing it has. My hypothesis would be supported if there is a positive relationship between public boat access and zebra mussel presence. A negative relationship would disprove this, which would be if a rise in zebra mussels was shown without public boat access.

```{r}
library(tidyverse)
library(readxl)
library(dplyr)
```
```{r}
Boat_Access<-read_csv("Data/Public_Boat_Access_Sites.csv")
AIS<-read_csv("Data/WatersWithAIS (1).csv")
Boat_Access
AIS
```

```{r}
Boat_Access<- Boat_Access |>
  rename(`Waterbody Name`= `WATERBODY_NAME_TEXT`)
```
```{r}
Boat_Access$lake_name <- toupper(Boat_Access$`Waterbody Name`)
AIS$lake_name <- toupper(AIS$`Waterbody Name`)

zebra <- AIS |>
  filter("Invasive Species" == "Zebra Mussel")

zebra_presence <- zebra |>
  distinct(`Waterbody Name`) |>
  mutate(zm_present = 1)
```
```{r}
lake_data <- Boat_Access |>
  left_join(zebra_presence, by = "Waterbody Name")
lake_data <- lake_data |>
  mutate(zm_present = ifelse(is.na(zm_present), 0, 1))
lake_data
```
```{r}
lake_data<- lake_data |> select(`Waterbody Name`, LANDING_TYPE_CODE, RAMP_TYPE_CODE, zm_present)
lake_data
```
## need to put in values that are not containing public landings 
